`clarify`: Simulation-Based Inference for Regression Models

### 2025-09-18 rjournal_paper/Final version/clarify.Rmd

---
title: "`clarify`: Simulation-Based Inference for Regression Models"
date: "`r Sys.Date()`"
abstract: >
  Simulation-based inference is an alternative to the delta method for computing the uncertainty around regression post-estimation (i.e., derived) quantities such as average marginal effects, average adjusted predictions, and other functions of model parameters. It works by drawing model parameters from their joint distribution and estimating quantities of interest from each set of simulated values, which form a simulated "posterior" distribution of the quantity from which confidence intervals can be computed. `clarify` provides a simple, unified interface for performing simulation-based inference for any user-specified derived quantities as well as wrappers for common quantities of interest. `clarify` supports a large and growing number of models through its interface with the `marginaleffects` package and provides native support for multiply imputed data.
draft: true
author:  
  - name: Noah Greifer
    affiliation: Harvard University
    address:
    - Institute for Quantitative Social Science
    - Cambridge, MA
    url: ngreifer.github.io
    orcid: 0000-0003-3067-7154
    email:  ngreifer@iq.harvard.edu
  - name: Steven Worthington
    affiliation: Harvard University
    address:
    - Institute for Quantitative Social Science
    - Cambridge, MA
    email: sworthington@iq.harvard.edu
    orcid: 0000-0001-9550-5797
  - name: Stefano Iacus
    affiliation: Harvard University
    address:
    - Institute for Quantitative Social Science
    - Cambridge, MA
    email: siacus@iq.harvard.edu
    orcid: 0000-0002-4884-0047
  - name: Gary King
    affiliation: Harvard University
    address:
    - Institute for Quantitative Social Science
    - Cambridge, MA
    email: king@harvard.edu
    orcid: 0000-0002-5327-7631
type: package
output: 
  rjtools::rjournal_pdf_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
---

@@

More formally, we fit a regression model $y_i = f(x_i; \beta)$, such as a linear or other generalized linear model with model coefficients $\beta$. We assume$$
\hat{\beta} \sim \text{MVN}(\beta, \Sigma_{\hat{\beta}})
$$where $\hat{\beta}$ is the vector of estimates of $\beta$ and $\Sigma_{\hat{\beta}}$ is their asymptotic covariance matrix. We define a function $\tau(\beta)$ that represents a quantity of interest derived from the model parameters, and compute its estimate $\widehat{\tau(\beta)}$ as $\tau(\hat{\beta})$. To perform simulation-based inference, we take $M$ draws $\tilde{\beta}^{(j)}$ for $j\in(1, \dots, M)$ from a multivariate Normal distribution with mean vector $\mu = \hat{\beta}$ and covariance $\Sigma = \hat{\Sigma}_{\hat{\beta}}$, where $\hat{\Sigma}_{\hat{\beta}}$ is an estimate of the asymptotic covariance matrix of the parameter estimates. We use the distribution of $\tilde{\tau}=\tau(\tilde{\beta})$ as the "posterior" distribution of $\widehat{\tau(\beta)}$, and compute its variance as$$
\hat\sigma^2_{\widehat{\tau(\beta)}} = \frac{1}{M-1}\sum^M_{j=1}{(\tilde{\tau}^{(j)} - \bar{\tilde{\tau}})^2}
$$and quantile $100(1-\alpha)\%$ confidence interval limits as $\left[\tilde{\tau}_{(\frac{\alpha}{2})}, \tilde{\tau}_{(1-\frac{\alpha}{2})}\right]$ where $\tilde{\tau}_{(q)}$ is the $q$th value of $\tilde{\tau}$ when arranged in ascending order (i.e., the $q$th quantile of the empirical cumulative distribution function). Simulation-based Wald-type confidence intervals can be computed as $\left[\tau(\hat{\beta}) + \hat\sigma_{\widehat{\tau(\beta)}} Z_{\frac{\alpha}{2}}, \tau(\hat{\beta}) + \hat\sigma_{\widehat{\tau(\beta)}} Z_{1-\frac{\alpha}{2}} \right]$, where $Z_q$ is the $q$th quantile of a standard Normal distribution. The delta method-based Wald-type confidence intervals use this formula but with the first-order Taylor approximation to the asymptotic variance: $\hat\sigma^2_{\widehat{\tau(\beta)}}=\nabla\tau(\hat{\beta}) \Sigma_{\hat{\beta}} \nabla\tau'(\hat{\beta})$, where $\nabla\tau(\hat{\beta})$ is the gradient of $\tau(\beta)$ with respect to $\beta$ evaluated at $\hat{\beta}$.

To compute a p-value for a hypothesis test involving the quantity of interest, i.e., $\text{H}_0: \tau (\beta) = \tau_0$ with a given null value $\tau_0$, we can invert the confidence interval [@thulinModernStatisticsWrangling2021]; that is, we find the largest value of $\alpha$ such that $\tau_0$ is within the confidence interval and use that $\alpha$ as the p-value for the test. For Wald-based inference (either using the simulation-based variance or delta method-based variance), this is equivalent to performing a standard two-sided Z-test using the test statistic $Z=\left(\tau(\hat\beta)-\tau_0\right) / \hat\sigma_{\widehat{\tau(\beta)}}$. One benefit of using the quantile p-values for inference is that equivalent tests of the same hypothesis will always yield identical p-values; for example, testing the equality of two derived quantities will yield the same p-value when comparing the difference between the quantities against a null hypothesis of 0 and the ratio of the quantities against a null hypothesis of 1, as each of these hypothesis is true if and only if the other is true.

@@

# Related Software

Similar functionality exists in the \pkg{CLARIFY} package in Stata[^2] [@tomzClarifySoftwareInterpreting2003] and used to be available in the `Zelig` R package [@imaiCommonFrameworkStatistical2008a], though there are differences in these implementations. \CRANpkg{clarify} provides additional flexibility by allowing the user to request any derived quantity, in addition to providing shortcuts for common quantities, including predictions at representative values, average marginal effects, and average dose-response functions (described below). \CRANpkg{clarify} relies on and can be seen as a companion to the \CRANpkg{marginaleffects} package [@marginaleffectsforthcoming], which offers similar functionality but primarily uses the delta method for calculating uncertainty (though simulation-based inference is supported in a more limited capacity as well).

@@

-   `method` -- if `ci` is `TRUE`, the method used to compute the bounds. Allowable methods include thea Normal approximation (`"wald"`) or using the quantiles of the resulting distribution (`"quantile"`). The Normal approximation involves multiplying the standard deviation of the estimates (i.e., which functions like the standard error of the sampling distribution) by the critical Z-statistic computed using `(1-level)/2` to create a symmetric margin of error around the point estimate. The default is `"quantile"` to instead use quantile-based bounds.

@@

```{r plot1, fig.width=4}
plot(est1, reference = TRUE, ci = FALSE)
```

@@

-   `level` -- the desired two-sided confidence level. The default is .95 so that that the bounds are at the .025 and .975 quantiles when method (see below) is `"quantile"`.

@@

```{r plot2, fig.width=4}
plot(est2, reference = TRUE, ci = FALSE)

@@

# `sim_apply()` Wrappers: `sim_setx()`, `sim_ame()`, `sim_adrf()`

@@

`sim_setx()` provides an interface to compute predictions at representative and user-supplied values of the predictors. For example, we might want to know what the effect of treatment is for a "typical" individual, which corresponds to the contrast between two model-based predictions (i.e., one under treatment and one under control for a unit with "typical" covariate values). This functionality mirrors the `setx()` and `setx1()` functionality of `Zelig` (which is where its name originates) and provides similar functionality to functions in \CRANpkg{modelbased}, \CRANpkg{emmeans}, \CRANpkg{effects}, and \CRANpkg{ggeffects}.

@@

```{r plot3, fig.width=5}
plot(est3, var = "re75", ci = FALSE)
```

@@

```{r plot4, fig.width=5}
plot(est4)
```

@@

```{r plot8, fig.width=5}
plot(est8)
```

@@

```{r plot9, fig.width=5}
plot(est9)
```

@@

# Transforming and Combining Estimates

@@

# Using \pkg{clarify} with Multiply Imputed Data

@@

# Comparison to Other Packages

Several packages offer methods for computing interpretable quantities form regression models, including \CRANpkg{emmeans} [@emmeans2024], \pkg{margins} [@margins2021], \CRANpkg{modelbased} [@modelbased2020], and \CRANpkg{marginaleffects} [@marginaleffectsforthcoming]. Many of the quantities computed by these packages can also be computed by \CRANpkg{clarify}, the primary difference being that \CRANpkg{clarify} uses simulation-based inference rather than delta method-based inference.

>>>>

---
title: "`clarify`: Simulation-Based Inference for Regression Models"
date: "2025-02-11"
abstract: >
  Simulation-based inference is an alternative to the delta method for computing the uncertainty around regression post-estimation (i.e., derived) quantities such as average marginal effects, average adjusted predictions, and other functions of model parameters. It works by drawing model parameters from their joint distribution and estimating quantities of interest from each set of simulated values, which form a simulated "posterior" distribution of the quantity from which confidence intervals can be computed. clarify provides a simple, unified interface for performing simulation-based inference for any user-specified derived quantities as well as wrappers for common quantities of interest. clarify supports a large and growing number of models through its interface with the marginaleffects package and provides native support for multiply imputed data.
draft: true
author:  
  - name: Noah Greifer
    affiliation: Harvard University
    address:
    - Institute for Quantitative Social Science
    - Cambridge, MA
    url: ngreifer.github.io
    orcid: 0000-0003-3067-7154
    email:  ngreifer@iq.harvard.edu
  - name: Steven Worthington
    affiliation: Harvard University
    address:
    - Institute for Quantitative Social Science
    - Cambridge, MA
    email: sworthington@iq.harvard.edu
    orcid: 0000-0001-9550-5797
  - name: Stefano Iacus
    affiliation: Harvard University
    address:
    - Institute for Quantitative Social Science
    - Cambridge, MA
    email: siacus@iq.harvard.edu
    orcid: 0000-0002-4884-0047
  - name: Gary King
    affiliation: Harvard University
    address:
    - Institute for Quantitative Social Science
    - Cambridge, MA
    email: king@harvard.edu
    orcid: 0000-0002-5327-7631
type: package
output: 
  rjtools::rjournal_pdf_article:
    toc: no
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
---

@@

More formally, we fit a regression model $y_i = f(x_i; \beta)$, such as a linear or other generalized linear model with model coefficients $\beta$. We assume $\hat{\beta} \sim \text{MVN}(\beta, \Sigma_{\hat{\beta}})$ where $\hat{\beta}$ is the vector of estimates of $\beta$ and $\Sigma_{\hat{\beta}}$ is their asymptotic covariance matrix. We define a function $\tau(\beta)$ that represents a quantity of interest derived from the model parameters, and compute its estimate $\widehat{\tau(\beta)}$ as $\tau(\hat{\beta})$.

To perform simulation-based inference, we take $M$ draws $\tilde{\beta}^{(j)}$ for $j\in(1, \dots, M)$ from a multivariate Normal distribution with mean vector $\mu = \hat{\beta}$ and covariance $\Sigma = \hat{\Sigma}_{\hat{\beta}}$, where $\hat{\Sigma}_{\hat{\beta}}$ is an estimate of the asymptotic covariance matrix of the parameter estimates. We use the distribution of $\tilde{\tau}=\tau(\tilde{\beta})$ as the "posterior" distribution of $\widehat{\tau(\beta)}$, and compute its variance as
\begin{equation}
\hat\sigma^2_{\widehat{\tau(\beta)}} = \frac{1}{M-1}\sum^M_{j=1}{(\tilde{\tau}^{(j)} - \bar{\tilde{\tau}})^2}
(\#eq:posteriorvariance)
\end{equation} 
and quantile $100(1-\alpha)\%$ confidence interval limits as $\left[\tilde{\tau}_{(\frac{\alpha}{2})}, \tilde{\tau}_{(1-\frac{\alpha}{2})}\right]$ where $\tilde{\tau}_{(q)}$ is the $q$th value of $\tilde{\tau}$ when arranged in ascending order (i.e., the $q$th quantile of the empirical cumulative distribution function of $\tilde{\tau}$). Simulation-based Wald-type confidence intervals can be computed as
\begin{equation}
\left[\tau(\hat{\beta}) + \hat\sigma_{\widehat{\tau(\beta)}} Z_{\frac{\alpha}{2}}, \tau(\hat{\beta}) + \hat\sigma_{\widehat{\tau(\beta)}} Z_{1-\frac{\alpha}{2}} \right]
(\#eq:siminterval)
\end{equation}
where $Z_q$ is the $q$th quantile of a standard Normal distribution. The delta method-based Wald-type confidence intervals use this formula but with the first-order Taylor approximation to the asymptotic variance:
\begin{equation}
\hat\sigma^2_{\widehat{\tau(\beta)}}=\nabla\tau(\hat{\beta}) \Sigma_{\hat{\beta}} \nabla\tau'(\hat{\beta})
(\#eq:deltavariance)
\end{equation}
where $\nabla\tau(\hat{\beta})$ is the gradient of $\tau(\beta)$ with respect to $\beta$ evaluated at $\hat{\beta}$.

To compute a p-value for a hypothesis test involving the quantity of interest, i.e., $\text{H}_0: \tau (\beta) = \tau_0$ with a given null value $\tau_0$, we can invert the confidence interval [@thulinModernStatisticsWrangling2021]; that is, we find the largest value of $\alpha$ such that $\tau_0$ is within the confidence interval and use that $\alpha$ as the p-value for the test. For Wald-based inference (either using the simulation-based variance or delta method-based variance), this is equivalent to performing a standard two-sided Z-test using the test statistic
\begin{equation}
Z=\frac{\tau(\hat\beta)-\tau_0}{\hat\sigma_{\widehat{\tau(\beta)}}}
(\#eq:waldz)
\end{equation}
One benefit of using the quantile p-values for inference is that equivalent tests of the same hypothesis will always yield identical p-values; for example, testing the equality of two derived quantities will yield the same p-value when comparing the difference between the quantities against a null hypothesis of 0 and the ratio of the quantities against a null hypothesis of 1, as each of these hypothesis is true if and only if the other is true.

@@

# Related software

Similar functionality exists in the \pkg{CLARIFY} package in Stata[^2] [@tomzClarifySoftwareInterpreting2003] and used to be available in the \pkg{Zelig} R package [@imaiCommonFrameworkStatistical2008a], though there are differences in these implementations. \CRANpkg{clarify} provides additional flexibility by allowing the user to request any derived quantity, in addition to providing shortcuts for common quantities, including predictions at representative values, average marginal effects, and average dose-response functions (described below). \CRANpkg{clarify} relies on and can be seen as a companion to the \CRANpkg{marginaleffects} package [@marginaleffectsforthcoming], which offers similar functionality but primarily uses the delta method for calculating uncertainty (though simulation-based inference is supported in a more limited capacity as well).

@@

-   `method` -- if `ci` is `TRUE`, the method used to compute the bounds. Allowable methods include the Normal approximation (`"wald"`) or using the quantiles of the resulting distribution (`"quantile"`). The Normal approximation involves multiplying the standard deviation of the estimates (i.e., which functions like the standard error of the sampling distribution) by the critical Z-statistic computed using `(1-level)/2` to create a symmetric margin of error around the point estimate. The default is `"quantile"` to instead use quantile-based bounds.

@@

```{r plot1, fig.width=4, fig.alt="A plot of the density of the simulated predicted probability for participant PSID1, overlaid with a Normal distribution and a reference line for the median of the estimates"}
plot(est1, reference = TRUE, ci = FALSE)
```

@@

-   `level` -- the desired two-sided confidence level. The default is .95 so that that the bounds are at the .025 and .975 quantiles when `method` (see below) is `"quantile"`.

@@

```{r plot2, fig.width=4, fig.alt = "A plot of the density of the simulated difference between two regression coefficients, overlaid with a Normal distribution and a reference line for the median of the estimates."}
plot(est2, reference = TRUE, ci = FALSE)

@@

# Wrappers for `sim_apply()`: `sim_setx()`, `sim_ame()`, and `sim_adrf()`

@@

`sim_setx()` provides an interface to compute predictions at representative and user-supplied values of the predictors. For example, we might want to know what the effect of treatment is for a "typical" individual, which corresponds to the contrast between two model-based predictions (i.e., one under treatment and one under control for a unit with "typical" covariate values). This functionality mirrors the `setx()` and `setx1()` functionality of \pkg{Zelig} (which is where its name originates) and provides similar functionality to functions in \CRANpkg{modelbased}, \CRANpkg{emmeans}, \CRANpkg{effects}, and \CRANpkg{ggeffects}.

@@

```{r plot3, fig.width=5, fig.alt = "Two plots of the densities of the simulated predictions of the outcome for control and treated units, one corresponding to re74 = 0 and the other to re75 = 20000."}
plot(est3, var = "re75", ci = FALSE)
```

@@

```{r plot4, fig.width=5, fig.alt = "A plot of the predicted values of the outcome across levels of re75 from 0 to 20000 with a line for control units and a line for treated units. Each line has an uncertainty band around it."}
plot(est4)
```

@@

```{r plot8, fig.width=5, fig.alt = "A plot of the average dose-response function for the outcome against age. The line has an uncertainty band around it."}
plot(est8)
```

@@

```{r plot9, fig.width=5, fig.alt = "A plot of the average marginal effect function for the outcome against age. The line has an uncertainty band around it."}
plot(est9)
```

@@

# Transforming and combining estimates

@@

# Using \CRANpkg{clarify} with multiply imputed data

@@

# Comparison to other packages

Several packages offer methods for computing interpretable quantities form regression models, including \CRANpkg{emmeans} [@emmeans2024], \CRANpkg{margins} [@margins2021], \CRANpkg{modelbased} [@modelbased2020], and \CRANpkg{marginaleffects} [@marginaleffectsforthcoming]. Many of the quantities computed by these packages can also be computed by \CRANpkg{clarify}, the primary difference being that \CRANpkg{clarify} uses simulation-based inference rather than delta method-based inference.
